---
title: "Final Project: Chicago Divvy Ridership"
output:
  pdf_document: default
---



$\\$



## Author: Swathi Nachiar Manivannan & Freddy Gonzalez

## Discussants: https://www.kaggle.com/datasets/yingwurenjian/chicago-divvy-bicycle-sharing-data








<!--  


This is a template for creating your final project report. It lays out the
sections that should be in your write-up and describes a little about these
sections. There is some flexibility to deviate from this structure, for example,
interweaving more visualizations and analyses could work well.

Your report should be between 5-8 pages long and should contain:

    1. Introduction: 
      a. What is question you are addressing? 
      b. Why is important? 
      c. Where did you get the data?
      d. What other analyses that been done on the data ?
      
    2. Visualizations of the data: one or more plots
    
    3. Analyses: models, hypothesis tests, confidence intervals and other
    inferential statistics that give insight into your question
    
    4. Conclusions: What you found, future directions, etc.
    
    5. Reflection (do be completed on Canvas)
       a. What went well? 
       b. What did you struggle with?
       c. What analyses did you do that you are not including? etc. 

Please make your report look good by paying attention to detail, using
additional R Markdown features etc.

If there is additional code or data you would like to include with your report,
please create a GitHub page and add a link to it in your report. Additionally,
you can append the full code for your analysis in an appendix section at the end
of the document, and then include only the most important pieces of code in the
body of the report. For example, you can exclude details of data cleaning from
the body of the report. However, include anything central to your analyses, and
also any information about particular choices you made that might affect the
results, and why you made those choices, in the body of the report (e.g.,
explain if data was excluded and why, etc.).



--> 


<!-- There are some options that might help make your document look better.  
Feel free to add additional options here -->
```{r message=FALSE, warning=FALSE, tidy=TRUE, echo=FALSE}

library(knitr)

# This makes sure the code is wrapped to fit when it creates a pdf
opts_chunk$set(tidy.opts=list(width.cutoff=60))   


# Set the random number generator to always give the same random numbers
set.seed(230)  


```







$\\$









## Introduction 


<!--  

Write ~1-3 paragraphs describing:

1. What is problem you are addressing and why the problem you are addressing is
interesting.

2. Where you got the data from, including a link to the website where you got
the data if applicable.

3. What other analyses have already been done with the data and possibly links
to other analyses. Also mention if you are using the data in another class or
for another research project.


--> 

Bike-sharing systems have been increasingly introduced in cities around the world 
to offer residents an alternative and more affordable mode of transportation. Particularly 
in light of increased traffic congestion and anthropogenically caused climate change and 
pollution, there has been a greater need to offer "greener" modes of transportation in 
cities as a solution to these issues, and more importantly, incentivise residents to consider 
these options instead of driving. Bike-sharing systems have been met with mixed success all 
over the world; while some cities have had successful bike-sharing programs for several years, 
others have scrapped them within months from their launch. Therefore, in order to ensure the 
success of bike-sharing systems, there is first and foremost a need to characterise the 
ridership so that sensible and sustainbale financial decisions can 
be made based on this information.

Divvy is a bike-sharing system that services the Greater Chicago metropolitan area. It was 
first introduced in 2013 and has since become an integral part of Chicago, with an annual 
ridership of more than 3 million as of 2019. The aim of our project is to therefore 
characterise and obtain a broad picture understanding of Divvy ridership in the year 2017. 
The main questions we are asking are:

1. What is the geographical distribution of Divvy bike rides in the Greater Chicago metropolitan area?
2. How does the mean/frequency of bike ridership vary with respect to:
  a. Gender of commuters
  b. Temporality?
    i. The hour of the day?
    ii. The day of the week?
    iii. The month of the year?
  c. Seasonality?
    i. Temperature
    ii. Precipitation events?
3. Can we construct a model to predict the duration and/or distance of Divvy bike trips?


Possible analyses:
- calculate distance based on latitude-longitude data and use that as a response variable/predictor variable?
- what is the main response variable we want to determine? 
  - duration of trips? 
  - distance of trips?
  - busiest stations?
- potential predictor variables
  - gender
  - month
  - weather ("events" and "temperature")



   
$\\$   
   
    




## Results



$\\$
    


### Data wrangling: Change the subtitle here to describe what you are plotting etc.



<!--  

Very briefly discuss how you got the data into shape for your analyses. You can
include some code here, although extensive data cleaning code should be put on
GitHub and/or in an appendix at the end of the document.

--> 




$\\$
    




### Visualize the data: Change the subtitle here to describe what you are plotting etc.



<!--  

Create one or more plots of your data. Describe the plot(s), what they shows,
and why they are of interest to your analysis. Include the code to create these
plots in the R chunk below. You can also discuss the plots after the code too.
Finally, be sure to make plots as clear as possible (clear axis labeling,
legends and captions) so that it is easy for the reader to quickly understand
the central information being conveyed.

--> 
Data Visualizations 1.1 - Geographical Visualization

The Divvy dataset provides the bike station where the trip starts and ends. The 
location of the street is defined in terms of street name as well as in terms of 
latitude/longitude. We can use this information to locate the stations on a map 
and to study the trip behavior for Divvy users in the greater Chicago metro area. 
This plot is of particular interest because it will help elucidate the differences 
in ridership under different influencing variables. 

For this specific graph, we will be observing the geographical distribution of 
Divvy bike rides with respect to rider gender

The first thing we are going to do is try to have a geographical representation
of where in the city most of the trips start. This can be done using the ggmap
package within R. The pipeline that we will follow is such that: 

1. We will download a map of the area we are interested in, and then overlay on
the map an additional layer in the form of a contour plot. The script below 
specifies for a map of Chicago centered around the loop, a 35-block area of 
downtown Chicago, and any neighborhoods immediately adjacent including the 
West Loop and Gold Coast, for example. 

2. We then plot the map with ggmap() and finally add the contour plot that 
represents the frequency of trips originated at each Divvy bike station. Darker 
areas represent locations in which many trips start, while lighter colors 
represent the opposite. 




```{r message=FALSE, warning=FALSE, tidy=TRUE}

# read in csv containing divvy bike data. This is the raw data file obtained from Kaggle

divy_chicago <- read.csv("data.csv")

# filter to only include ridership data in 2017 and omit any NA

divy_2017 <- divy_chicago |>
  filter(year == "2017") |>
  na.omit()



library(ggplot2)
library(ggmap)

# register Google API key

register_google(key = "AIzaSyC5NF4xccTOw2UvjV4PRIElzrX8vr5Yv6U")

# load Chicago map and make sure it loads properly
# also adjust zoom scale to only visualize loop and surrounding areas 

chicago <- get_map(location = "Chicago", zoom = 13, legend = "bottom")

chicago.map <- ggmap(chicago)

# use lon and lat data to indicate start and end points that will be used to make
# map

from <- data.frame(lon = as.numeric(divy_2017$longitude_start), lat = as.numeric(divy_2017$latitude_start))

to <- data.frame(lon = as.numeric(divy_2017$longitude_end), lat = as.numeric(divy_2017$latitude_end))

# A) create a map that shows most common starting points by gender 

from1 <- data.frame(from, gender = divy_2017$gender)

chicago.map + stat_density2d(aes(x=lon, y=lat, fill = ..level..), alpha=0.5, 
size = 2, bins = 8, data=from1, geom="polygon") + scale_fill_gradient(low="pink", high="purple4") + facet_wrap(~ gender) + ggtitle("Ridership By Gender") + 
  xlab("Longitude") + ylab("Latitude") 

```

Discussion of graph 1.1A - Ridership by gender 

This graph generates a contour plot for men and women separately. Male riders 
seem to cover a larger a larger area of starting points, albeit less predictable. 
On the other end, female riders seem to have a more predictable starting location, 
generating more concentrated starting points at fewer locations. However, there
is a large overlap among both genders at locations near Fulton Market, the 
Merchandise Mart, and near Clark/Lake. 

<!--  

Possible additional discussion of the plots here. 

--> 

Data Visualizations 1.2 - Differences in ridership by day of the week  

The Divvy dataset provides the day of the week in which a trip was taken. 
According to map B, plotting ridership by day of the week, there are differences in where 
trips were taken from throughout the week. This poses the question of how time of day
may influence ridership throughout the week. This plot is of particular interest because it 
will help elucidate which time points are most popular among users. 

For this specific graph, we will be observing the distribution of ridership with 
respect to hour and day of the week. The first thing we are going to do is is 
briefly summarize ridership by usertype. There are two main types of Divvy riders, 
subscribers and customers. Subscribers pay $99/yr and get unlimited 
rides so long as they are under 45 minutes. These are more than often commuters 
that live in the city and travel to and from work. Customers pay as they go. This is 
a great option for tourists to see Chicago for cheap. 

The first thing we are going to do is briefly summarize ridership by usertype: 

1. We first changed the user type data in our dataset to be contained as a factor. 
We then created a summary table to count ridership by usertype. 

2. Our summary table indicated that Subscribers were the most common riders during 
2017, with a total of 2956854 rides. Because of this, we chose to merely focus 
on this user type. 

To visualize Subscriber ridership by hour throughout the week, our pipeline consisted 
of the following steps: 

1. Subset for both 'Subscriber' user type and day of the week. We chose to split 
this up into two plots, one looking at Subscriber ridership throughout the work
week and another looking at Subscriber ridership on the weekends.

2. Plot this subset of data in accordance with hour and percentage of rides at 
that hour. 


```{r message=FALSE, warning=FALSE, tidy=TRUE}

# count riders by usertype to gain inference on who is most utilizing the Divvy
# bike service 

divy_2017$usertype <- as.factor(divy_2017$usertype)

summary(divy_2017$usertype)

# subset for subscribers 

divy_count_hour_2017 <- divy_2017 %>%
  select(usertype, month, week, day, hour, tripduration) %>%
  group_by(hour,day) %>%
  subset(usertype == "Subscriber") %>%
  na.omit()

# subset for subscribers and exclude weekends to rule out tourists 

divy_count_hour_2017 %>% subset( !(day %in% c("5", "6"))) %>%
ggplot(aes(x = hour, fill = day)) + geom_bar(position = "dodge") + 
  scale_fill_brewer(palette = "YlOrRd") + 
  labs(y = "Counts", x = "Hour", fill = "Day") + 
  ggtitle("Figure 1.2A - Subscriber Ridership Throughout Work Week")

# now let's take a look at weekend ridership

divy_count_hour_2017 %>% subset( !(day %in% c("0","1","2","3","4" ))) %>%
ggplot(aes(x = hour, fill = day)) + geom_bar(position = "dodge") + 
  scale_fill_brewer(palette = "YlOrRd") + 
  labs(y = "Counts", x = "Hour", fill = "Day") + 
  ggtitle("Figure 1.2B - Subscriber Ridership on Weekends")


# divy_count_hour_2017$day <- c("Sunday",
                       #  "Monday",
                       #  "Tuesday",
                        # "Wednesday", 
                        # "Thursday", 
                        # "Friday", 
                       #  "Saturday" 
                       # )

```

Discussion of graph 1.2 - Differences in proportions of trips taken by Subscribers
throughout the week

This graph generates a plot for ridership differences among 'Subscribers' by hour 
throughout the week, and is separated into two plots:

Plot 1.2A graphs 'Subscriber' ridership throughout the work week (Mon-Fri). This 
plot shows two peaks at which riders are more often using Divvy. Those two peaks 
occur during the morning and afternoon rush hours when riders are traveling to 
work and traveling back from work, respectively. Ridership is at a minimum during 
the very early hours when most riders are still asleep. 

Plot 1.2B graphs 'Subscriber' ridership throughout the weekend (Sat-Sun). This plot 
shows a different pattern in which riders are utilizing Divvy throughout most of the
day. An elongated peak can be seen between hours ~10-19 when most riders are enjoying 
their weekend. While rides are still at a minimum during the very early hours of the day, 
there is a higher proportion of riders at these hours as opposed to those same hours 
on throughout the work week. This is likely due to riders staying out later on the 
weekends. 

$\\$    
    
Data Visualizations 1.3 - Trip Duration Throughout Week (Draft)

Now that we've plotted how many trips are being taken throughout the week, we 
want to take a look at how long these trips (in minutes) generally are throughout 
the week. We're again going to focus on Subscribers, since they are the most observed 
population throughout our data set. This visualization will identify what day(s)
of the week riders spend most of their time utilizing Divvy,allowing us to further 
characterize when there is a greater demand the bikes. 

For this visualization, the following pipeline was used: 

1. Use dplyr to create a new data frame containing variables such as user type, 
hour, day, and trip duration. From here, we grouped the data frame using hour and 
day, subsetted for Subscribers, and omitted any NA variables. 

2. We then used this data frame to plot the ride time (in minutes) using a heatmap.


```{r message=FALSE, warning=FALSE, tidy=TRUE}

# summary stat for df defining average ride time

mean(divy_count_hour_2017$tripduration)

# do a whole plot of trip duration and subset for Subscribers 
divy_count_hour_2017 <- divy_2017 %>%
  select(usertype, month, week, day, hour, gender, tripduration, temperature, events, trip_id) %>%
  group_by(hour, day) %>%
  subset(usertype == "Subscriber") %>%
  na.omit()

# heat map of trip duration

ggplot(divy_count_hour_2017, aes(hour, day, fill = tripduration)) + geom_tile() +
  guides(fill = guide_colourbar(barwidth = 1, barheight = 6.5)) + 
  coord_fixed() + labs(x = "Hour", y = "Day", fill = "Trip Duration (min)") + 
  ggtitle("Figure 1.3A - Divvy Bike Trip Duration by Hour of the Day")

# histogram to show distribution of trip duration among Subscribers 

ggplot(divy_count_hour_2017, aes(tripduration, fill = day)) + 
  geom_histogram(aes(tripduration), bins = 100, binwidth = 0.1) + 
  labs(x = "Trip Duration", y = "Counts", fill = "Day") + 
  ggtitle("Figure 1.3B - Trip Duration Distribution Among Subscribers")

# df for trips taken
divy_count_trips_taken <- divy_2017 %>%
  select(usertype, month, week, day, hour, gender, tripduration, temperature, events, trip_id) %>%
  group_by(day, week) %>%
  subset(usertype == "Subscriber") %>%
  count(usertype = "Subsciber") %>%
  na.omit()

# heatmap of trips taken
ggplot(divy_count_trips_taken, aes(week, day, fill = n)) + geom_tile() +
  guides(fill = guide_colourbar(barwidth = 1, barheight = 7)) + 
  coord_fixed() + labs(x = "Week", y = "Day", fill = "Trips Taken") + 
  ggtitle("Figure 1.3C - Divvy Trips Taken Throughot the Year") 

```

Discussion of Visualization 1.3 - Trip Duration Throughout Week


    

### Analyses: Sub-title about the analyses/models you are using 



<!--  

Build linear models, run hypothesis tests, create confidence intervals and/or
run simulations to answer questions that are of interest.

--> 

Analyses 1.3 - Anova for gender differences in ridership 

First need to check conditions for Anova 

1. We can check the first condition, as to whether the variances within each group
are approximately the same, by comparing the standard deviations of ride durations
within each group (gender). As long as the largest standard deviation is
not twice as big as the smallest standard deviation this condition is roughly
met.


```{r message=FALSE, warning=FALSE, tidy=TRUE}

library(dplyr)

# first subset for Subscribers then visualize to see if there might be any obvious 
# differences 

divy_subscribers <- divy_2017 %>%
  select(usertype, month, week, day, hour, gender, tripduration, trip_id) %>%
  group_by(gender) %>%
  subset(usertype == "Subscriber") %>%
  na.omit()

divy_subscribers %>%
  ggplot(aes(tripduration, gender, fill = gender)) +
  geom_boxplot() + coord_flip() + ggtitle("Trip Duration vs Gender") + 
  labs(x = "Gender", y = "Trip Duration (min)", fill = "Gender")

# CONDITION 1

# first subset for Subscribers 

divy_subscribers <- divy_2017 %>%
  select(usertype, month, week, day, hour, gender, tripduration, trip_id) %>%
  group_by(gender) %>%
  subset(usertype == "Subscriber") %>%
  na.omit()

(obs_stat <- anova(lm(tripduration ~ gender, data = divy_subscribers))$F[1])

# checking this condition with dplyr
gender_stats <- divy_subscribers |>
  group_by(gender) |>
  summarize(mean_tripduration = mean(tripduration),
            sd_reaction = sd(tripduration))


# print table 
gender_stats

# check ratio between largest and smallest sd (CONDITION 1 MET)
# ensure that the ratio is not twice as big between the smallest and largest deviation 
gender_stats$sd_reaction
max(gender_stats$sd_reaction)/min(gender_stats$sd_reaction)

# CONDITION 2 

# create residual plot 

gender_residuals <- divy_subscribers %>%
  group_by(gender) %>%
  mutate(mean_rt = mean(tripduration), 
            rt_residuals = tripduration - mean_rt)


qqnorm(gender_residuals$rt_residuals)

# normalize 

gender_residuals_log <- divy_subscribers %>%
  group_by(gender) %>%
  mutate(log_tripduration = (log(tripduration)))


gender_residuals <- gender_residuals_log %>%
  group_by(gender) %>%
  mutate(mean_log_rt = mean(log_tripduration), 
            rt_residuals = log_tripduration - mean_log_rt)


qqnorm(gender_residuals$rt_residuals)

# let's now create our anova table 

(fit_tripduration <- lm(tripduration ~ gender, data = divy_subscribers))

(anova_table <- anova(fit_tripduration))

# plot null distribution

(obs_stat <- anova(lm(log_tripduration ~ gender, data = gender_residuals_log))$F[1])

# this is as far as i got here
# define df1, df2, df3

(N <- nrow(gender_residuals))
(K <- 2)

(df1 <- K - 1)
(df2 <- N - K)
(df3 <- N -1)


# visualize the null distribution 
x <- seq(-.5, 5, by = 0.01)
y <- df(x, df1, df2)

plot(x, y, type = "l", xlim = c(0,10))
abline(v = obs_stat, col = "red")

```

Condition 1
The variances (or standard deviations) in each group must be approximately the same 

-the ratio between the smallest and largest sd is not twice as big, so this
condition seems to be met. 






$\\$


    
    
    
    
    
    


## Conclusion 



<!--  


~1-2 paragraphs summarizing what you found, how the findings address your
question of interest, and possible future directions. Please make sure describe
your conclusions in an intuitive way, and make sure that your argument is strong
and backed by solid evidence from your data.



-->










$\\$






## Reflection


<!--  


Reflection  

Write one paragraph describing what went well with this project and what was
more difficult. Also describe any additional things you tried that you did not
end up including in this write-up, and approximately how much time you spend
working the project.

Finally, please go to Canvas and answer a few questions related to how this project went: https://yale.instructure.com/courses/79947/quizzes/52421



-->




$\\$




## Appendix


<!--  


You can include a complete listing of your code here if you could not fit it
into the body of the document. Make sure your code is well commented and easy to
read - i.e., use meaningful object names, separate your code into sections,
describe what each section is doing, use good formatting, etc.


-->




