---
title: "Final Project: Add title here"
output:
  pdf_document: default
---



$\\$



## Author:  Freddy Gonzalez

## Discussants: 
https://www.kaggle.com/datasets/yingwurenjian/chicago-divvy-bicycle-sharing-data

https://www.kaggle.com/datasets/laxmena/divvy-ridesharing-dataset 






<!--  


This is a template for creating your final project report. It lays out the
sections that should be in your write-up and describes a little about these
sections. There is some flexibility to deviate from this structure, for example,
interweaving more visualizations and analyses could work well.

Your report should be between 5-8 pages long and should contain:

    1. Introduction: 
      a. What is question you are addressing? 
      b. Why is important? 
      c. Where did you get the data?
      d. What other analyses that been done on the data ?
      
    2. Visualizations of the data: one or more plots
    
    3. Analyses: models, hypothesis tests, confidence intervals and other
    inferential statistics that give insight into your question
    
    4. Conclusions: What you found, future directions, etc.
    
    5. Reflection (do be completed on Canvas)
       a. What went well? 
       b. What did you struggle with?
       c. What analyses did you do that you are not including? etc. 

Please make your report look good by paying attention to detail, using
additional R Markdown features etc.

If there is additional code or data you would like to include with your report,
please create a GitHub page and add a link to it in your report. Additionally,
you can append the full code for your analysis in an appendix section at the end
of the document, and then include only the most important pieces of code in the
body of the report. For example, you can exclude details of data cleaning from
the body of the report. However, include anything central to your analyses, and
also any information about particular choices you made that might affect the
results, and why you made those choices, in the body of the report (e.g.,
explain if data was excluded and why, etc.).



--> 









<!-- There are some options that might help make your document look better.  
Feel free to add additional options here -->
```{r message=FALSE, warning=FALSE, tidy=TRUE, echo=FALSE}

library(knitr)

# This makes sure the code is wrapped to fit when it creates a pdf
opts_chunk$set(tidy.opts=list(width.cutoff=60))   


# Set the random number generator to always give the same random numbers
set.seed(230)  


```







$\\$









## Introduction 


<!--  

1. Introduction: 
      a. What is question you are addressing? 
      b. Why is important? 
      c. Where did you get the data?
      d. What other analyses that been done on the data ?

Write ~1-3 paragraphs describing:

1. What is the problem you are addressing and why the problem you are addressing is
interesting.

Divvy is a bike-sharing system in the Greater Chicago 

Possible analyses:
- calculate distance based on latitude-longitude data and use that as a response variable/predictor variable?
- what is the main response variable we want to determine? 
  - duration of trips? 
  - distance of trips?
  - busiest stations?
- potential predictor variables
  - gender
  - month
  - weather ("events" and "temperature")
  - 



2. Where you got the data from, including a link to the website where you got
the data if applicable.

I got the dataset from Kaggle, which takes a look at Divvy bike usership in Chicago from 2014 to 2017

https://www.kaggle.com/datasets/sudalairajkumar/daily-temperature-of-major-cities


3. What other analyses have already been done with the data and possibly links
to other analyses. Also mention if you are using the data in another class or
for another research project.

I have not previously done any analyses with the data, but a quick google search 
yields prior analyses with similar data such as that in the link listed below. 
I am not using this data for another project or class. 

https://www.weather.gov/btv/climate 

--> 





   
$\\$   
   
    




## Results

Analyses: models, hypothesis tests, confidence intervals and other
    inferential statistics that give insight into your question


$\\$
    


### Data wrangling: Change the subtitle here to describe what you are plotting etc.

Data wrangling 1.1 - Data read-in and initial sort 

This section reads in Divvy bike use data from Chicago. Additional data sorting 
contained within this section. 

```{r initial_data_cleanup}

library(dplyr)
library(doBy)
library(readr)
library(ggplot2)

# read in csv containing divvy bike data 

divy_chicago <- read.csv("data.csv")

# filter to only include ridership data in 2017 and omit any NA

divy_2017 <- divy_chicago %>%
  filter(year == "2017") %>%
  na.omit()

# create date variable 

divy_2017$date <- as.Date(divy_2017$starttime)

divy_chicago$date <- as.Date(divy_chicago$starttime)

# create count variable by date and gender

divy_count_2017 <- divy_2017 %>% count(date, gender)

divy_count <- divy_chicago %>% count(date, gender)

# create count by hour and gender 
# this could be used to look at trip fluctuation during rush hours 
# could also look at just weekends and see how they compare 

divy_count_hour_2017 <- divy_2017 %>%
  count(hour, gender)

```

Data wrangling 1.2 - Data summaries 

This section contains summary statistics that will allow for broad overview of 
trends that will be visualized in later sections. 

``` {r summary_statistics}


# summarize average temperature by month 

summaryBy(temperature ~ month, divy_2017 , FUN = mean)

# initial analysis on usership by gender 

summary(divy_2017$gender)

divy_2017$gender <- as.factor(divy_2017$gender)

summary(divy_2017$gender)

summaryBy(tripduration ~ gender, divy_2017, FUN = mean)

# multivariate summary trip duration by month and gender 

summaryBy(tripduration ~ month*gender, divy_2017, FUN = mean)

# usertype data analysis 

mean(divy_2017$tripduration) 

divy_2017$usertype   <- as.factor(divy_2017$usertype)

summary(divy_2017$usertype)

# weather summary 

divy_2017$events <- as.factor(divy_2017$events)

summary(divy_2017$events)


```


Data wrangling 1.3a - Visualizations 

This section contains some initial visualizations using base R. 

``` {r base_r_visualizations}

# boxplot of avg temp by month 

boxplot(temperature ~ month, data = divy_2017, 
        main = "average temperatures during bike ridership")

divy_2017 %>%
  ggplot(aes(temperature, month, fill = month)) + 
  geom_boxplot() + 
  xlab("Month") + 
  ylab("Temperature")



```


Data wrangling 1.3b - Visualizations 

This section contains further visualizations using ggplot. 

Figure out which, if any, of these to keep for final submission. 

``` {r ggplot_visualizations}

# gender data analysis 

ggplot(divy_2017) + geom_bar(aes(gender, y = (..count..)/sum(..count..)), fill="pink") + ylab("Proportion") + theme_bw() + ggtitle("ridership proportion by gender")


# histogram of trip duration

ggplot(divy_2017) + geom_histogram(aes(tripduration)) + xlab("trip duration (minutes)") + 
  ggtitle("trip duration histogram for chicago ridership in 2017")

# density curves

# density curve of trip duration by gender
divy_2017 %>% ggplot()  + 
  geom_density(aes(tripduration,group=gender, fill=gender, colour=gender), adjust=3, alpha=0.1) +
  ggtitle("differences in ridership duration by gender")

# density curve showing difference in ridership by gender throughout year
divy_2017 %>% ggplot()  + 
  geom_density(aes(month,group=gender, fill=gender, colour=gender), adjust=3, alpha=0.1) + 
  ggtitle("ridership differences by gender throughout the year")

# density curve showing difference in ridership by gender in differing temperatures
divy_2017 %>% ggplot()  + 
  geom_density(aes(temperature,group=gender, fill=gender, colour=gender), adjust=3, alpha=0.1) + 
  ggtitle("ridership differences by gender in differing temperatures")


# ridership differences throughout the week by gender 
divy_2017 %>% ggplot()  + 
  geom_density(aes(day,group=gender, fill=gender, colour=gender), adjust=3, alpha=0.1) + 
  ggtitle("ridership differences by gender in differing temperatures")

# trip duration mean by gender 

ggplot(divy_2017) + geom_boxplot(aes(tripduration, group = gender, colour = gender))


# plotting proportions of usertypes by hour 
ggplot(divy_2017) + 
  geom_bar(aes(x=hour, y=(..count..)/sum(..count..), fill=usertype)) + 
  theme_bw() + ylab("counts") + ggtitle("usertype count by hour")

# subset for subscribers and exclude weekends to rule out tourists 

divy_2017 %>% subset( (usertype == "Subscriber") & !(day %in% c("5","6"))) %>%
ggplot(.) + geom_bar(aes(x=hour, y=(..count..)/sum(..count..), fill=day), position = "dodge") +  ylab("") + scale_fill_brewer(palette = "YlOrRd")

# plot of trip counts by date for 2017
divy_count_2017 %>%
  ggplot(aes(date, n, col = gender)) + 
  geom_jitter(position = position_jitter(height = .2)) + 
  xlab("date") + ylab("# of trips") + 
  ggtitle("trips by gender")

divy_count_2017 %>%
  ggplot(aes(date, n, col = gender)) + 
  geom_smooth() + xlab("date") + ylab("# of trips") + ggtitle("trips by gender")

# plot of trip counts by date for ~5 years
divy_count %>% 
  ggplot(aes(date, n, col = gender)) + 
  geom_jitter(alpha = .4, position = position_jitter(height = .2)) + 
  xlab("date") + ylab("# of trips") + 
  ggtitle("trips by gender")

divy_count %>%
  ggplot(aes(date, n, col = gender)) + 
  geom_smooth() + xlab("date") + ylab("# of trips") + ggtitle("trips by gender")




```

```{r clean_data}


# time series of trip duration as a function of temperature 

plot(divy_2017$tripduration, divy_2017$temperature, type = "o")


# create chicago map 

library(ggmap)

# registration

register_google(key = "AIzaSyC5NF4xccTOw2UvjV4PRIElzrX8vr5Yv6U")

# create map 
chicago <- get_map(location = "Chicago", zoom = 13, legend = "bottom")

chicago.map <- ggmap(chicago)

from <- data.frame(lon = as.numeric(divy_2017$longitude_start), lat = as.numeric(divy_2017$latitude_start))

to <- data.frame(lon = as.numeric(divy_2017$longitude_end), lat = as.numeric(divy_2017$latitude_end))

from1 <- data.frame(from, gender = divy_2017$gender)

chicago.map + stat_density2d(aes(x=lon, y=lat, fill = ..level..), alpha=0.5, 
size = 2, bins = 8, data=from1, geom="polygon") + scale_fill_gradient(low="pink", high="purple4") + facet_grid(~ gender) #+ theme(panel.margin = unit(2, "lines"))




```

<!--  

Very briefly discuss how you got the data into shape for your analyses. You can
include some code here, although extensive data cleaning code should be put on
GitHub and/or in an appendix at the end of the document.

--> 




$\\$
    




### Visualize the data: Change the subtitle here to describe what you are plotting etc.

Visualizations of the data: one or more plots

<!--  

Create one or more plots of your data. Describe the plot(s), what they shows,
and why they are of interest to your analysis. Include the code to create these
plots in the R chunk below. You can also discuss the plots after the code too.
Finally, be sure to make plots as clear as possible (clear axis labeling,
legends and captions) so that it is easy for the reader to quickly understand
the central information being conveyed.

--> 




```{r message=FALSE, warning=FALSE, tidy=TRUE}
library(ggplot2)

# further filter to only include data on nyc and san diego 



```




<!--  

Possible additional discussion of the plots here. 

--> 








$\\$    
    







    

### Analyses: Sub-title about the analyses/models you are using 



<!--  

Build linear models, run hypothesis tests, create confidence intervals and/or
run simulations to answer questions that are of interest.

--> 




```{r message=FALSE, warning=FALSE, tidy=TRUE}




```








$\\$


    
    
    
    
    
    


## Conclusion 



<!--  


~1-2 paragraphs summarizing what you found, how the findings address your
question of interest, and possible future directions. Please make sure describe
your conclusions in an intuitive way, and make sure that your argument is strong
and backed by solid evidence from your data.



-->










$\\$






## Reflection


<!--  


Reflection  

Write one paragraph describing what went well with this project and what was
more difficult. Also describe any additional things you tried that you did not
end up including in this write-up, and approximately how much time you spend
working the project.

Finally, please go to Canvas and answer a few questions related to how this project went: https://yale.instructure.com/courses/79947/quizzes/52421



-->




$\\$




## Appendix


<!--  


You can include a complete listing of your code here if you could not fit it
into the body of the document. Make sure your code is well commented and easy to
read - i.e., use meaningful object names, separate your code into sections,
describe what each section is doing, use good formatting, etc.


-->




